Local Spectral Clustering of Density Upper Level Sets

### 2019-05-21 neurips/local_spectral_density_clustering_tuesday.tex
added Tuesday edits for neurips draft

\usepackage{neurips_2019}

@@

\newcommand{\diam}{\mathrm{diam}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\cut}{\mathrm{cut}}
% \newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\1}{\mathbf{1}}
\newcommand{\Phibf}{\mathbf{\Phi}}
\newcommand{\Psibf}{\mathbf{\Psi}}
\newcommand{\dist}{\mathrm{dist}}

@@

To be more precise, let $G=(V,E,w)$ denote a weighted, undirected graph  
constructed from the samples $\Xbf$, where $V=\{1,\ldots,n\}$, and $w_{uv}
= K(x_u,x_v) \geq 0$ for $u,v \in V$, and a particular kernel function $K$.
Here $(u,v) \in E$ if and only if $w_{uv} > 0$.  We denote by $\Abf \in
\Reals^{n \times n}$ the weighted adjacency matrix, which has entries
$A_{uv}=w_{uv}$, and by $\Dbf$ the degree matrix, with 
\smash{$\Dbf_{uu} = \sum_{v \in V} \Abf_{uv}$}.  We also denote by $\Wbf,\Lbf$
the (lazy) random walk transition probability matrix and normalized\footnote{Other
	popular choices here include the unnormalized Laplacian, and symmetric
	normalized Laplacian.} 
Laplacian matrix, respectively, which are defined as
$$
\Wbf = \frac{\Ibf{} + \Dbf^{-1}\Abf}{2}, \quad \Lbf = \Ibf{} - \Wbf,
$$
where $\Ibf{} \in \Reals^{n\times n}$ is the identity matrix.  Classical global
spectral methods take a eigendecomposition $L=U \Sigma U^T$, use some 
number of eigenvectors (columns in $U$) as a feature representation for the
samples, and then run (say) k-means in this new feature space.

When applied to geometric graphs constructed from a large number of samples,
global spectral clustering methods can be computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to recent increased interest in
local spectral algorithms, which leverage locally-biased spectra computed using
random walks around a user-specified seed node.  A popular local clustering
algorithm is Personalized PageRank (PPR), first introduced by
\citep{haveliwala2003}, and further developed by
\citep{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013},
among others.

Local spectral clustering techinques have been practically very successful
\citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, which has led
many authors to develop supporting theory
\citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case
guarantees on traditional graph-theoretic notions of cluster quality (like
conductance).  In this paper, we adopt a more traditional statistical viewpoint,
and examine what the output of a local clustering algorithm on $\Xbf$ reveals
about the unknown density $f$.  In particular, we examine the ability of the PPR
algorithm to recover \emph{density clusters} of $f$, which are defined as the
connected components of the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$
for some threshold $\lambda > 0$ (a central object of central interest in the
classical statistical literature on clustering, dating back to
\citet{hartigan1981}).

\subsection{PPR on a Neighborhood Graph}

We now describe the clustering algorithm that will be our focus for the rest of 
the paper. We start with the geometric graph that we form based on the samples 
$\Xbf$: for a radius $r > 0$, we consider the \emph{$r$-neighborhood graph} of 
$\Xbf$, denoted $G_{n,r}=(V,E)$, an unweighted graph with vertices
$V=\Xbf$, and an edge $(x_i,x_j) \in E$ if and only if $\norm{x_i - x_j}
\leq  r$, where $\norm{\cdot}$ denotes Euclidean norm.  Note that this is a
special case of the general construction introduced above, with 
$K(u,v) = 1(\norm{x_u - x_v} \leq r)$.

Next, we define the PPR vector $\pbf = \pbf(v,\alpha;G_{n,r})$, with respect to  
a seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to be
the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf$ is the random walk matrix of the underlying graph $G_{n,r}$ 
and $e_{v}$ denotes indicator vector for node $v$ (with a 1 in the $v$th
position and 0 elsewhere).  In practice, we can approximately solve the above
linear system via a simple, efficient random walk, with appropriate restarts to
$v$.

For a level $\beta > 0$ and a target volume $\vol_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf = (p_u)_{u \in V}$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \{u \in V: \frac{p_u}{\Dbf_{uu}} > \frac{\beta}{\vol_{0}}\}.
\end{equation}

Having computed sweep cuts $S_{\beta}$ over a range \smash{$\beta \in  
(\frac{1}{40},\frac{1}{11})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{1}{40}, \frac{1}{11})$} is standard in the analysis of PPR
algorithms, see, e.g., \citep{zhu2013}.}, we then  output a cluster estimate $\widehat{C} = S_{\beta^*}$ to have minimum normalized cut $\Phi(S_{\beta^{\star}}; G_{n,r})$, where for $S \cup S^c = G_{n,r}$, $\cut(S;G_{n,r}) := |\set{(u,v) \in E : u \in S, v \in S^c}|$, $\vol(S; G_{n,r}) := \sum_{u \in S} \Dbf_{uu}$, and 
\begin{equation}
\label{eqn: normalized_cut}
\Phi(S; G_{n,r}) := \frac{\cut(S;G_{n,r})}{\min \set{\vol(S; G_{n,r}), \vol(S^c; G_{n,r})}}.
\end{equation}
For concreteness, we summarize this procedure in Algorithm \ref{alg: ppr}.

@@

\subsection{Summary of Results}

% It is worth calling attention to some other work on computing the normalized
% cut over neighborhood graphs. In this context, continuous analogues to (for
% instance) normalized cut have been defined, over the data-manifold rather than
% the graph, and convergence finite sample graph-theoretic functionals to their
% continuous counterparts has been shown 
% \cite{garciatrillos16, arias-castro12, maier11}. 
% However, in addition to the graph-minimization problem being computationally
% infeasible, these continuous analogues are not always easily interpretable --
% and their corresponding minimizers not always easily identifiable -- for the
% particular density function under consideration. Of course, relating these
% partitions to the arguably more simply defined high density clusters can be
% also challenging in general. Intuitively, however, under the right conditions
% such high-density clusters should have more edges within themselves than to
% the remainder of the graph. We formalize this intuition next.

%% RJT: I didn't know where to put this.  It was out of place, and now I'm not
%% sure where it goes ... should it go in related work?

%It is worth pointing out that in this context, some theory has been developed
%regarding how graph theoretic quantities such as the normalized cut $\Phi$ (and
%others) relate to properties of the underlying distribution $f$ as well as the
%kernel function $k$. Such analyses typically proceed by defining a continuous
%analogue to the measure of cluster quality under consideration. Then, under
%appropriate specification of $k$ and a proper schedule of $(r_n)$, convergence
%of clusters output by spectral (and other) algorithms to the corresponding
%minima of these continuous analogues has been shown \cite{vonluxburg2008,
%garciatrillos18}.

%% RJT: This was already commented out before

Let $\Cbb_f(\lambda)$ denote the connected components of the density upper level
set $\{x \in \Rd: f(x) > \lambda\}$.  For a given density cluster $\Cset \in
\Cbb_f(\lambda)$, we call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical
	density cluster}. Below we give two notions of performance of a density cluster estimate.

\begin{definition}[Misclassification error]
	\label{def: misclassification_rate}
	For an estimator \smash{$\Cest \subseteq \Xbf$} and set
	$\mathcal{S} \subseteq \Reals^d$, the \emph{misclassification error} of $\mathcal{S}$ by $\Cest$ is
	\begin{equation}
	\label{eqn: misclassification_rate}
	\abs{\Cest \setminus (\mathcal{S} \cap \Xbf)} + \abs{(\mathcal{S} \cap \Xbf) \setminus \Cest}.
	\end{equation}
\end{definition}

\begin{definition}[Consistent density cluster estimation]
	\label{def: consistent_density_cluster_estimation}
	For an estimator \smash{$\Cest \subseteq \Xbf$} and cluster 
	$\Cset \in \Cbb_f(\lambda)$, we say \smash{$\Cest$} is a consistent
	estimator of $\Cset$ if for all $\Cset' \in \Cbb_f(\lambda)$ with $\Cset \not=
	\Cset'$ the following holds as $n \to \infty$: 
	\begin{equation}
	\label{eqn: consistent_density_cluster_recovery}
	\Cset[\Xbf] \subseteq \Cest \quad \text{and} \quad
	\Cest \cap \Cset'[\Xbf] = \emptyset,
	\end{equation}
	with probability tending to 1.
\end{definition}

A summary of our main results (and outline for the rest of this paper) is as
follows.

\begin{enumerate}
	\item In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we introduce a set of natural geometric conditions. We formalize a measure of difficulty based on these geometric conditions, and show that when properly initialized, the misclassification error of Algorithm \ref{alg: ppr} is upper bounded by this difficulty measure.

	\item We further show that if the density cluster $\Cset$ is particularly well-conditioned, Algorithm \ref{alg: ppr} will perform consistent density cluster estimation in the sense of \eqref{eqn: consistent_density_cluster_recovery}.

	\item Corollary \ref{cor: appr} establishes that these statements hold also with respect to an approximate form of \ppr, which can be efficiently computed.

	\item In \textcolor{red}{Section 3}, we detail some of the main technical machinery required to prove our main results, highlighting the part various geometric quantities play in the ultimate difficulty of the clustering problem.

@@

On the topic of conditions, it is worth mentioning that, as density clusters
are inherently local, focusing on the PPR algorithm actually eases our analysis
and allows us to require fewer global regularity conditions relative to those
needed for more classical global spectral algorithms.

\subsection{Related Work}
%In addition to the background given above, a few related lines of work are worth
%highlighting. For neighborhood graphs of the type we consider, continuous analogues to (for
%instance) normalized cut have been defined, over the data-manifold rather than
%the graph, and convergence finite sample graph-theoretic functionals to their
% continuous counterparts has been shown 
% \cite{garciatrillos16, arias-castro12, maier11}.

In addition to the background given above, a few related lines of work are worth
highlighting. Building on earlier work of
\citep{koltchinskii2000}, \citep{vonluxburg2008,hein2005} studied the limiting behaviour of spectral clustering
algorithms. These authors show that when samples are obtained from a
distribution, and we appropriately construct a geometric graph, the spectrum of
the Laplacian converges to that of the Laplace-Beltrami operator on the
data-manifold. However, relating the partition obtained using the
Laplace-Beltrami operator to the more intuitively defined high-density
clusters can be challenging in general.

% AJG 4/29: I need to rewrite to explain how its similar, because
% this is a good opportunity to show that implications
% of the type given in our work are broadly of interest.

More similar to our results are the works
\citep{vempala2004,shi2009,schiebinger2015}, who study the consistency of
spectral algorithms in recovering the latent labels in certain parametric and
nonparametric mixture models. These results focus on global rather than local
algorithms, and as such impose global rather than local conditions on the nature
of the density. Moreover, they do not in general ensure recovery of density
clusters, which is the focus in our work.

@@

\subsection{Geometric Conditions on Density Clusters}
\label{subsec: geometric_conditions}

As mentioned previously, successful recovery of a density cluster by \pprspace requires the density cluster to be geometrically well-conditioned. At a minimum, we wish to avoid dumbbell-like sets $\Cset$ which contain an arbitrarily thin bridge, and as in \cite{chaudhuri2010} we therefore introduce a buffer zone around $\Cset$. Letting $B(x,r)$ be the closed ball of radius $r > 0$ centered at $x \in \Rd$, for a given cluster $\Cset \subseteq \Rd$ and $\sigma > 0$, we refer to $\Csig := \set{y \in \Reals^d: \inf_{x \in \Cset} \norm{y - x} \leq \sigma}$ as the $\sigma$-expansion of $\Cset$, and state our conditions with respect to $\Csig$.

More generally, over the neighborhood graph $G_{n,r}$ we would like the empirical cluster $\Csig[\Xbf]$ to be \textcolor{red}{well connected} everywhere in its interior, and \textcolor{red}{poorly connected} to the rest of $\Xbf$.  This intuition motivates our required conditions, stated with respect to a density
cluster $\Cset \in \Cbb_f(\lambda)$ for some threshold $\lambda > 0$, and an
expansion parameter $\sigma > 0$.

\begin{enumerate}[label=(A\arabic*)]
	\item
	\label{asmp: bounded_density}
	\emph{Bounded density within cluster:} There are $0 < \lambda_{\sigma} <
	\Lambda_{\sigma} < \infty$ such that
	$$
	\lambda_{\sigma} = \inf_{x \in \Csig} f(x) \leq \sup_{x \in \Csig} f(x) \leq
	\Lambda_{\sigma}.
	$$
	% and 
	% \begin{equation*}
	% \frac{\diam \Asig}{\sigma} \leq \mu
	% \end{equation*}
	% where $\diam \Asig = \sup \set{d(x,y) : x,y \in \Asig}$

	\item
	\label{asmp: cluster_separation}
	\emph{Cluster separation:}
	For all $\Cset' \in \Cbb_f(\lambda)$ with $\Cset' \not= \Cset$,
	$$
	\dist(\Csig,\Csig') > \sigma,
	$$
	where \smash{$\dist(\Cset,\Cset') = \inf_{x \in \Cset} \dist(x,\Cset')$}.

	\item 
	\label{asmp: low_noise_density}
	\emph{Low noise density:} There exists $\gamma,c_0 > 0$ such that for all $x
	\in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
	$$
	\inf_{x' \in \Csig} f(x') - f(x) \geq  c_0 \dist(x, \Csig)^{\gamma},
	$$
	where \smash{$\dist(x,\Cset) = \inf_{x_0 \in \Cset} \norm{x - x_0}$}.

	%%% AJG 5/20: Should I turn this into two conditions?
	\item
	\label{asmp: embedding}
	\emph{Lipschitz embedding:}
	$\Csig$ is the image of a convex set under a biLipschitz, measure preserving mapping. Formally, there exists $\mathcal{K} \subseteq \Rd$ convex, and $g: \Reals^d \to \Reals^d$ such that $\det(\nabla g (x)) = 1$ for all $x \in \Csig$, and for some $L \geq 1$,
	\begin{equation*}
	\frac{1}{L}\norm{x - y} \leq \norm{g(x) - g(y)} \leq L \norm{x - y} ~ \text{for all $x,y \in \Csig$}
	\end{equation*}
	such that $\Csig$ is the image of $\mathcal{K}$ by $g$, $\Csig = g(\mathcal{K})$.
	Furthermore, there exists $D < \infty$ such that for all $x, x' \in \mathcal{K}$
	$$
	\norm{x - x'} \leq D.
	$$
	\item
	\label{asmp: bounded_volume}
	\emph{Bounded volume:}
	Let the neighborhood graph radius $0 < r \leq \sigma/2d$ be such that
	\begin{equation}
	\label{eqn: weighted_cluster_volume}
	\frac{\int_{\Csig} \Pbb(B(x,r)) f(x) dx }{\int_{\Rd} \Pbb(B(x,r)) f(x) dx} \leq \frac{1}{2}.
	\end{equation}
\end{enumerate}

The cluster separation \ref{asmp: cluster_separation} and low noise density \ref{asmp: low_noise_density} conditions guarantee \textcolor{red}{poor connectivity} between $\Csig[\Xbf]$ and $\Xbf \setminus \Csig[\Xbf]$, whereas \ref{asmp: bounded_density} and \ref{asmp: embedding} ensure high connectivity within $\Csig[\Xbf]$. \textcolor{red}{It may not be immediately obvious how \ref{asmp: embedding} contributes to geometric conditioning. For now, we observe merely that random walks will mix slowly over sets with large diameter, and make some more detailed commentary in Section 3.} Finally, \ref{asmp: bounded_volume} is a relatively harmless technical condition, merely excluding the case where $\Csig$ contains over half the total mass.

\subsection{Well-Conditioned Density Clusters}

We turn to formally defining a \textcolor{red}{condition number}, $\kappa(\Cset)$, reflects the difficulty of the local spectral clustering task. The smaller $\kappa(\Cset)$ is, the more success \pprspace will have in recovering $\Cset$. Let $\theta := (r, \sigma, \lambda, \lambda_{\sigma}, \Lambda_{\sigma}, \gamma, D, L)$ contain those geometric parameters detailed in \ref{subsec: geometric_conditions}.

\begin{definition}[Well-conditioned density clusters]
	For $\lambda > 0$ and $\Cset \in \Cbb_f(\lambda)$, let $\Cset$ satisfy \ref{asmp: bounded_density} - \ref{asmp: bounded_volume} for some $\theta$, and \textcolor{red}{ additionally let $\Csig$ satisfy \eqref{eqn: weighted_cluster_volume} }. Then, setting
	\begin{align}
	\label{eqn: condition_number_1}
	\mathbf{\Phi}(\theta) 
	& := c_1 r \frac{d}{\sigma} \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} - c_0 \frac{r^{\gamma}}{\gamma + 1})}{\lambda_{\sigma}} \nonumber \\
	\mathbf{\Psi}(\theta) & := \Biggl(c_2 \frac{\Lambda_{\sigma}^4 d^3 D^2 L^2}{\lambda_{\sigma}^4 r^2} \log^2\left(\frac{1}{r}\right) + c_3 \log\left(\frac{\Lambda_{\sigma}}{\lambda_{\sigma}}\right) \Biggr)^{-1}
	\end{align}
	and
	\begin{equation}
	\label{eqn: condition_number}
	\kappa(\Cset) := \frac{\Phibf(\theta)}{\Psibf(\theta)}
	\end{equation}
	we call $\Cset$ a \textrm{$\kappa$-well-conditioned density cluster}.
\end{definition}

>>>>

\PassOptionsToPackage{numbers}{natbib}
\usepackage{neurips_2019}

@@

\newcommand{\diam}{\mathrm{diam}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\cut}{\mathrm{cut}}
% \newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\1}{\mathbf{1}}
\newcommand{\Phibf}{\Phi_{u}}
\newcommand{\Psibf}{\Psi_{u}}
\newcommand{\dist}{\mathrm{dist}}

@@

When applied to geometric graphs constructed from a large number of samples,
global spectral clustering methods can be computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to recent increased interest in
local spectral algorithms, which leverage locally-biased spectra computed using
random walks around a user-specified seed node.  A popular local clustering
algorithm is Personalized PageRank (\ppr), first introduced by
\citet{haveliwala2003}, and further developed in
\citep{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013},
among others.

Local spectral clustering techinques have been practically very successful
\citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, which has led
many authors to develop supporting theory
\citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case
guarantees on traditional graph-theoretic notions of cluster quality (like
conductance).  In this paper, we adopt a more traditional statistical viewpoint,
and examine what the output of a local clustering algorithm on $\Xbf$ reveals
about the unknown density $f$.  In particular, we examine the ability of the \pprspace algorithm to recover \emph{density clusters} of $f$, which are defined as the
connected components of the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$
for some threshold $\lambda > 0$ (a central object of central interest in the
classical statistical literature on clustering, dating back to
\citet{hartigan1981}).

\paragraph{PPR on a Neighborhood Graph.}We now describe the clustering algorithm that will be our focus for the rest of 
the paper. We start with the geometric graph that we form based on the samples 
$\Xbf$: for a radius $r > 0$, we consider the \emph{$r$-neighborhood graph} of 
$\Xbf$, denoted $G_{n,r}=(V,E)$, an unweighted, undirected graph with vertices
$V=\Xbf$, and an edge $(x_i,x_j) \in E$ if and only if $\norm{x_i - x_j}
\leq  r$, where $\norm{\cdot}$ denotes Euclidean norm. We denote by $\Abf \in \Reals^{n \times n}$ the adjacency matrix, with entries $\Abf_{uv} = 1$ if and only if $(u,v) \in E$, and by $\Dbf$ the degree matrix, with $\Dbf_{uu} = \sum_{v \in V} \Abf_{uv}$.

Next, we define the PPR vector $\pbf = \pbf(v,\alpha;G_{n,r})$, with respect to  
a seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to be
the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf = (\Ibf{} + \Dbf^{-1}\Abf)/2$ is the (lazy) random walk matrix over $G_{n,r}$ 
and $e_{v}$ denotes indicator vector for node $v$ (with a 1 in the $v$th
position and 0 elsewhere).

For a level $\beta > 0$ and a target volume $\vol_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf = (p_u)_{u \in V}$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \set{u \in V: \frac{p_u}{\Dbf_{uu}} > \frac{\beta}{\vol_{0}}}.
\end{equation}
For a set $S \subseteq V$ with complement $S^c = V \setminus S$, the normalized cut is defined to be 
\begin{equation}
\label{eqn: normalized_cut}
\Phi(S; G_{n,r}) := \frac{\cut(S;G_{n,r})}{\min \set{\vol(S; G_{n,r}), \vol(S^c; G_{n,r})}}.
\end{equation}
where $\cut(S;G_{n,r}) = \sum_{u \in S, v \in S^c} \Abf_{uv}$ and $\vol(S; G_{n,r}) := \sum_{u \in S} \Dbf_{uu}$. Having computed sweep cuts $S_{\beta}$ over a range \smash{$\beta \in  
(\frac{1}{40},\frac{1}{11})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{1}{40}, \frac{1}{11})$} is standard in the analysis of PPR
algorithms, see, e.g., \citep{zhu2013}.}, we then output the cluster estimate $\widehat{C} = S_{\beta^*}$ which has minimum normalized cut $\Phi(S_{\beta^{\star}}; G_{n,r})$. For concreteness, we summarize this procedure in Algorithm \ref{alg: ppr}.

@@

\paragraph{Estimation of density clusters.}

Let $\Cbb_f(\lambda)$ denote the connected components of the density upper level
set $\{x \in \Rd: f(x) > \lambda\}$.  For a given density cluster $\Cset \in
\Cbb_f(\lambda)$, we call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical
density cluster}. The symmetric set difference between estimated and empirical cluster is perhaps the most frequently used metric to quantify cluster estimation error \citep{korostelev2012,polonik1995,rigollet2009}.

\begin{definition}[Symmetric set difference]
	\label{def: symmetric_set_diff}
	For an estimator \smash{$\Cest \subseteq \Xbf$} and set
	$\mathcal{S} \subseteq \Reals^d$, the symmetric set difference of $\Cest$ and $\mathcal{S}$ is 
	\begin{equation}
	\label{eqn: misclassification_rate}
	\Delta(\Cest, \mathcal{S}) = \abs{\Cest \setminus \mathcal{S}[\Xbf] \cup \mathcal{S}[\Xbf] \setminus \Cest}.
	\end{equation}
\end{definition}

However, the symmetric set difference does not account for the distance points in $\widehat{C} \setminus \mathcal{S}[\Xbf]$ may be from $\mathcal{S}$ \citep{singh2009}. Therefore we introduce a second notion of cluster estimation, first introduced by \citet{hartigan1981} and defined asymptotically, which measures whether $\widehat{C}$ can distinguish any two distinct elements $\Cset, \Cset' \in \Cbb_f(\lambda)$.

\begin{definition}[Consistent density cluster estimation]
	\label{def: consistent_density_cluster_estimation}
	For an estimator \smash{$\Cest \subseteq \Xbf$} and cluster 
	$\Cset \in \Cbb_f(\lambda)$, we say \smash{$\Cest$} is a consistent
	estimator of $\Cset$ if for all $\Cset' \in \Cbb_f(\lambda)$ with $\Cset \not=
	\Cset'$ the following holds as $n \to \infty$: 
	\begin{equation}
	\label{eqn: consistent_density_cluster_recovery}
	\Cset[\Xbf] \subseteq \Cest \quad \text{and} \quad
	\Cest \cap \Cset'[\Xbf] = \emptyset,
	\end{equation}
	with probability tending to 1.
\end{definition}
\paragraph{Summary of results.}
A summary of our main results (and outline for the rest of this paper) is as
follows.

\begin{enumerate}
	\item In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we introduce a set of natural geometric conditions, formalize a measure of difficulty based on these geometric conditions, and show that when properly initialized, the symmetric set difference of Algorithm \ref{alg: ppr} is upper bounded by this difficulty measure.

	\item We further show that if the density cluster $\Cset$ is particularly well-conditioned, Algorithm \ref{alg: ppr} will perform consistent density cluster estimation in the sense of \eqref{eqn: consistent_density_cluster_recovery}. In Corollary \ref{cor: appr}, we establish that both metrics of cluster estimation can be bounded with respect to an approximate form of \ppr, which can be efficiently computed.

	\item In Section \ref{sec: analysis}, we detail some of the main technical machinery required to prove our main results, and expose the part various geometric quantities play in the ultimate difficulty of the clustering problem.

@@

Our main takeaway can be summarized as follows: \ppr, run on a neighborhood graph, recovers geometrically compact high-density clusters.

\paragraph{Related Work. }

In addition to the background given earlier, a few related lines of work are worth
highlighting. Similar in spirit to our results are the works
\citep{shi2009,schiebinger2015}, who study the consistency of
spectral algorithms in recovering the latent labels in certain parametric and
nonparametric mixture models. These results focus on global rather than local
algorithms, and as such impose global rather than local conditions on the nature
of the density. Moreover, they do not in general ensure recovery of density
clusters, which is the focus in our work. Perhaps most importantly, these works rely on general cluster saliency conditions, which depend implicitly on many distinct geometric aspects of the cluster $\Cset$ under consideration. We make this dependence explicit, and in so doing, expose the role each geometric condition plays in the clustering problem.

Additionally, we note that density clustering and level set estimation is a well-studied problem. \citep{polonik1995, rigollet2009} study density clustering under symmetric set difference, \citep{tsybakov1997, singh2009} prove minimax optimal level set estimators under Hausdorff loss and \citep{hartigan1981, chaudhuri2010} consider consistent estimation of the cluster tree, to note but a few works on the subject. Our goal is not to improve on these results, or offer yet another algorithm for level set estimation; indeed, seen as a density clustering algorithm, \pprspace has none of the optimality guarantees of the previous works. This is fact a major point of our article: \pprspace can provably recover density clusters, but only under strong geometric conditions.

@@

We formalize some geometric conditions, before using these to define a measure $\kappa(\Cset)$ which encodes the difficulty \pprspace will have estimating $\Cset$. We motivate this measure, and the underlying geometric conditions, by giving density cluster estimation guarantees for Algorithm \ref{alg: ppr} in terms of $\kappa(\Cset)$.

\paragraph{Geometric Conditions on Density Clusters.}

As mentioned previously, successful recovery of a density cluster by \pprspace requires the density cluster to be geometrically well-conditioned. At a minimum, we wish to sets $\Cset$ which contain arbitrarily thin bridges or spikes, and therefore as in \cite{chaudhuri2010} we introduce a buffer zone around $\Cset$. Let $B(x,r)$ be the closed ball of radius $r > 0$ centered at $x \in \Rd$. For a some $\lambda > 0$, consider a given cluster $\Cset \in \Cbb_f(\lambda)$. We denote the distance between $x$ and $\Cset$ as $\dist(x,\Cset) := \inf_{y \in \Cset} \norm{y - x}$, and for a given $\sigma > 0$, we refer to $\Csig := \set{x \in \Reals^d: \dist(x,\Cset) \leq \sigma}$ as the $\sigma$-expansion of $\Cset$. We now state our conditions with respect to $\Csig$, and provide some intuition afterwards.

\begin{enumerate}[label=(A\arabic*)]
	\item
	\label{asmp: bounded_density}
	\emph{Bounded density within cluster:} There exist constants $\lambda_{\min}, \Lambda_{\min}$ $0 < \lambda_{\sigma} <
	\Lambda_{\sigma} < \infty$ such that $ \lambda_{\sigma} = \inf_{x \in \Csig} f(x) \leq \sup_{x \in \Csig} f(x) \leq \Lambda_{\sigma}.$

	\item
	\label{asmp: cluster_separation}
	\emph{Cluster separation:}
	For all $\Cset' \in \Cbb_f(\lambda)$ with $\Cset' \not= \Cset$, $\dist(\Csig,\Csig') > \sigma$,
	where \smash{$\dist(\Csig,\Csig') := \inf_{x \in \Csig} \dist(x,\Csig')$}.

	\item 
	\label{asmp: low_noise_density}
	\emph{Low noise density:} There exists $\gamma,c_0 > 0$ such that for all $x
	\in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
	$$
	\inf_{x' \in \Csig} f(x') - f(x) \geq  c_0 \dist(x, \Csig)^{\gamma},
	$$

	%%% AJG 5/20: Should I turn this into two conditions?
	\item
	\label{asmp: embedding}
	\emph{Lipschitz embedding:}
	There exists $g: \Reals^d \to \Reals^d$ which has the following properties: i) there exists a convex set $\mathcal{K} \subseteq \Rd$ with $\diam(\mathcal{K}) = \sup_{x,y \in \mathcal{K}}\norm{x - y} =: D < \infty$, such that $\Csig = g(\mathcal{K})$, ii) $\det(\nabla g (x)) = 1$ for all $x \in \Csig$, where $\nabla g(x)$ is the Jacobian of $g$ evaluated at $x$; iii) for some $L \geq 1$, 
	\begin{equation*}
	\frac{1}{L}\norm{x - y} \leq \norm{g(x) - g(y)} \leq L \norm{x - y} ~ \text{for all $x,y \in \mathcal{K}$}.
	\end{equation*}
	Simply put, $\Csig$ is the image of a convex set with finite diameter, under a biLipschitz, measure preserving transformation.
	\item
	\label{asmp: bounded_volume}
	\emph{Bounded volume:}
	Let the neighborhood graph radius $0 < r \leq \sigma/2d$ be such that
	\begin{equation}
	\label{eqn: weighted_cluster_volume}
	2 \int_{\Csig} \Pbb(B(x,r)) f(x) dx \leq \int_{\Rd} \Pbb(B(x,r)) f(x) dx
	\end{equation}
\end{enumerate}

Thinking of $\Csig[\Xbf]$ as a subset of vertices in $G_{n,r}$, we would like $\Csig[\Xbf]$ to be internally well-connected, while being poorly connected to the rest of $\Xbf$. The cluster separation \ref{asmp: cluster_separation} and low noise density \ref{asmp: low_noise_density} conditions guarantee low connectivity between $\Csig[\Xbf]$ and $\Xbf \setminus \Csig[\Xbf]$ in $G_{n,r}$, whereas \ref{asmp: bounded_density} and \ref{asmp: embedding} ensure high connectivity within $\Csig[\Xbf]$. It may not be immediately obvious how \ref{asmp: embedding} contributes to geometric conditioning. For now, we observe merely that random walks will mix slowly over sets with large diameter, and comment on this condition in more detail in Section \ref{sec: analysis}. Finally, \ref{asmp: bounded_volume} is a relatively harmless technical condition, merely excluding the case where $\vol(\Csig[\Xbf]; G_{n,r}) > \vol(\Xbf; G_{n,r})/2$.

We can now formally define the \textcolor{red}{condition number}, $\kappa(\Cset)$, which reflects the difficulty of the local spectral clustering task. The smaller $\kappa(\Cset)$ is, the more success \pprspace will have in recovering $\Cset$. Let $\theta := (r, \sigma, \lambda, \lambda_{\sigma}, \Lambda_{\sigma}, \gamma, D, L)$ contain those geometric parameters detailed in \ref{asmp: bounded_density} - \ref{asmp: bounded_volume}.

\begin{definition}[Well-conditioned density clusters]
	For $\lambda > 0$ and $\Cset \in \Cbb_f(\lambda)$, let $\Cset$ satisfy \ref{asmp: bounded_density} - \ref{asmp: bounded_volume} for some $\theta$, Then, for universal constants $c_1, c_2, c_3 > 0$ \textcolor{red}{to specified later}, we set
	\begin{equation}
	\label{eqn: condition_number_1}
	\Phibf(\theta) 
	:= c_1 r \frac{d}{\sigma} \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} - c_0 \frac{r^{\gamma}}{\gamma + 1})}{\lambda_{\sigma}},~ 
	\Psibf(\theta) := \Biggl(c_2 \frac{\Lambda_{\sigma}^4 d^3 D^2 L^2}{\lambda_{\sigma}^4 r^2} \log^2\left(\frac{1}{r}\right) + c_3 \log\left(\frac{\Lambda_{\sigma}}{\lambda_{\sigma}}\right) \Biggr)^{-1}
	\end{equation}
	and letting $\kappa(\Cset) := \frac{\Phibf(\theta)}{\Psibf(\theta)}$, we call $\Cset$ a \textrm{$\kappa$-well-conditioned density cluster}.
\end{definition}

At first glance \eqref{eqn: condition_number_1} may appear mysterious, but as will be shown in Section \ref{sec: analysis}, $\Phibf(\theta)$ and $\Psibf(\theta)$ are merely upper bounds on the normalized cut and inverse mixing time of $\Csig[\Xbf]$ in $G_{n,r}$. In \cite{zhu2013}, building on the work of \cite{andersen2006} and others, it is shown that the ratio of normalized cut to inverse mixing time is a fundamental quantity governing the clustering performance of \pprspace on a general graph. $\kappa(\Cset)$ upper bounds this ratio for an empirical density cluster over the neighborhood graph $G_{n,r}$, and is therefore a natural criterion to measure difficulty of the density clustering task.

\paragraph{Well-initialized algorithm.}
